A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2025-05-23 _tex/manuscript.tex
Built site for gh-pages

The datasets 1-4 are publicly available as CSV tables
\citep{ludwig_multi-centric_2023, ludwig_detailed_2022} and can be
interactively explored on \href{https://lyprox.org}{LyProX}. For each
patient the primary tumor subsite is reported and each individual LNL is
reported as either metastatic or healthy given the available diagnostic
modalities, which include pathology after neck dissection in some
patients. In this work we will stratify the tumor locations into
different ICD codes which are depicted in figure~\ref{fig-subsites}. For
glottis tumors (C32.0), only t-stages T2,T3 and T4 have been taken into
account as by definition glottis tumors are non-metastatic in the stages
T0 and T1.

@@

\caption{\label{fig-involvement}Prevalence of ipsilateral LNL
involvement stratified by subsite. The subsites are sorted in natural
order to represent the continuously changing LNL involvement. The
different tumo locations are color coded, where oral cavity subsistes
are depicted in blue, larynx in green, hypopharynx in red and larynx in
orange. The patient data is further stratified in early t-stage (0-2)
and late t-stage (3-4). The legend furter specifies the number of
patients in each subsite.}

@@

With equation~\ref{eq-marginalized-evolution} we can compute the
probability of a patient being in any state \(\xi_i\). Therefore the
likelihood of observing a the For model training we assume that the
diagnoses in our data \(\mathbf{D}\) we observe correspond to the hidden
state \(\mathbf{X}\) of the patient. Thus, learning the model parameters
corresponds to maximizing the probability of observing the dataset
\(\mathbf{D}\):

@@

In equation~\ref{eq-likelihood-HMM} we compute the likelihood of
observing patient the diagnosis of each patient \(k\), i.e.~t-stage
\(T_k\) and involvement \(\mathbf{X}_k=\boldsymbol{\xi}_i\).

@@

\begin{itemize}
\tightlist
\item
  \(\pi_m^{s_k}\) is the mixing coefficient for subsite \(s_k\) (where
  patient \(k\) has their tumor) and model \(m\),
\item
  \(P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)\) is the likelihood of
  patient \(k\)'s diagnosis, given that it was generated by model \(m\),
\item
  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
\item
  \(\epsilon_{k}^m\) is the latent variable that indicates patient \(k\)
  was generated by model \(m\).
\end{itemize}

@@

To actually find the mixing coefficients \(\boldsymbol{\pi}\) and all
models' parameters \(\boldsymbol{\theta}_m\) that maximize
equation~\ref{eq-incomplete-likelihood}, we follow an iterative approach
called \emph{expectation-maximization} or \emph{EM-algorithm}. With
arbitrarily initialized starting parameters, we alternate between the
following two steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In the \textbf{E}xpectation step, we compute the responsibilities
  \(\gamma(\epsilon_k^m)\), which represent the probabilities of a
  patient \(k\) originating from one of the models \(m\), given the
  current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
  as given in equation~\ref{eq-responsibilities}.
\item
  During the \textbf{M}maximization step, we find a new set of
  parameters that maximize the new expected complete data likelihood
  (equation~\ref{eq-expected-complete-likelihood}). For the mixture
  coefficients, we can even find an analytic solution to the new
  maximum: \[
  \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
  \] where we sum over the set of all patients \(K_s\) with their tumor
  in subsite \(s\).\\
  The models' new parameters are found by numerically maximizing the
  respective likelihood, weighted by the responsibilities:
  \begin{equation}\phantomsection\label{eq-maximization-step}{
  \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
  }\end{equation}
\end{enumerate}

@@

\section{Three component Mixture Model}\label{sec-3comp}

We illustrate the methodology for a mixture model with M = 3 components,
considering the ipsilateral involvement of LNLs I, II, III, IV, and V.
We include the ICD codes as subsites for oral cavity, hypopharynx and
oropharynx. In figure~\ref{fig-convergence} the convergence of the
negative log-likelihood and change in model parameters is depicted.
After a random inizialization, the algorithm rapidly converges. The
algorithm was stopped when the difference of log-likelihood between two
iterations was below 0.01.

@@

The spatial plot in figure~\ref{fig-3_simplex} illustrates how the model
assigns the three components to different tumor subsites. Component 0,
located at the bottom right of the triangle, primarily characterizes
oropharyngeal subsites. For instance, the base of tongue subsite (C01),
which exhibits the highest involvement of LNL II, is fully assigned to
this component. Similarly, subsite C10, which includes several
oropharyngeal regions, is assigned roughly 50\% to the oropharynx-like
component, with the remaining mixture distributed across the other two
components.

Hypopharyngeal subsites, on the other hand, are fully assigned to
Component 2, located at the top vertex of the triangle. Meanwhile, the
gum subsite (C03), with predominant LNL I involvement, is entirely
assigned to Component 1, situated at the bottom left. As subsites
anatomically approach the oropharynx, their mixture coefficients for the
oropharynx-like component increase. This is evident in the subsites C02
(tongue) and C05 (palate), which display a higher proportion of
oropharyngeal influence in their mixture. These results conform well
with the involvement patterns observed in the data.

@@

\includegraphics{figures/mixture_components_3_simplex_t_stages_new.png}

@@

\includegraphics{figures/mixture_components_3_t_stages_new.png}

@@

\section{Four component Mixture Model}\label{sec-4comp}

@@

The component assignment is shown in figure~\ref{fig-4-matrix}.
Similarly to the 3-component model the different tumor locations are
assigned to a one of the components\ldots.

Here i probably should permute the components such that we have the same
ordering as in the 3-component model.

@@

\includegraphics{figures/mixture_components_4_t_staging_add_larynx.png}

@@

\caption{\label{fig-4-matrix}Matrix representation of component
assignment. Each row of the matrix corresponds to each ICD code. The
collums represent the three different components}

@@

add some analysis (level specific predictions and also comparison to
single tumor location models.)

>>>>

Patients with glottic laryngeal cancer (C32.0) at tumor stages T0 and T1
have been excluded from analysis because these stages, by definition, do
not exhibit lymphatic involvement {[}need source from Panos{]}. The
datasets 1-4 are publicly available as CSV tables
\citep{ludwig_multi-centric_2023, ludwig_detailed_2022} and can be
interactively explored on \href{https://lyprox.org}{LyProX}. For each
patient the primary tumor subsite is reportd and each indicidual LNL is
reported as either metastatic or healthy given the available diagnostic
modalities, which include pathology after neck dissection in some
patients. In this work we will stratify the tumor locations into
different ICD codes which are depicted in figure~\ref{fig-subsites}.

@@

\caption{\label{fig-involvement}Prevalence of ipsilateral LNL
involvement stratified by subsite. The subsites are sorted in natural
order to represent the continuously changing LNL involvement. The
different tumo locations are color coded, where oral cavity subsistes
are depicted in blue, larynx in green, hypopharynx in red and larynx in
orange. The patient data is further stratified in early t-stage (0-2)
and late t-stage (3-4). The legend furter specifies the number of
patients in each subsite. For glottis (C32.0) early t-stage only
includes T2.}

@@

With equation~\ref{eq-marginalized-evolution} we can compute the
probability of a patient being in any state \(\xi_i\). To train the
model, we assume that the observed diagnoses in our dataset
\(\mathbf{D}\) directly reflect the underlying hidden states
\(\mathbf{X}\) of the patient. Thus, learning the model parameters
corresponds to maximizing the probability of observing the dataset
\(\mathbf{D}\):

@@

In equation~\ref{eq-likelihood-HMM} we compute the likelihood of
observing the diagnosis of each patient \(k\), i.e.~t-stage \(T_k\) and
involvement \(\mathbf{X}_k=\boldsymbol{\xi}_i\).

\subsection{Risk Estimation}\label{risk-estimation}

The goal of the HMM is to predict the risk of occult metastases in LNLs,
given a patient's diagnosis. In this framework this corresponds to
computing the probability of a hidden state
\(\mathbf{X}=\boldsymbol{\xi}_i\) given the observed diagnosis
\(\mathbf{Z}\) and the model parameters \(\boldsymbol{\theta}\). With
Baye's theorem we can compute this probability as follows:

\begin{equation}\phantomsection\label{eq-bayes}{
P(\mathbf{X} =\boldsymbol{\xi}_i \mid \mathbf{Z}, \boldsymbol{\theta}) = \frac{P(\mathbf{Z} \mid \mathbf{X}=\boldsymbol{\xi}_i, \boldsymbol{\theta}) P(\mathbf{X}_k=\boldsymbol{\xi}_i \mid \boldsymbol{\theta})}{P(\mathbf{Z} \mid \boldsymbol{\theta})}
}\end{equation}

where \(P(\mathbf{Z} \mid \boldsymbol{\theta})\) is the marginal
likelihood of the data, which can be computed by summing over all
possible hidden states \(\boldsymbol{\xi}_i\):

\begin{equation}\phantomsection\label{eq-marginal-likelihood}{
P(\mathbf{Z} \mid \boldsymbol{\theta}) = \sum_{\boldsymbol{\xi}_i} P(\mathbf{Z} \mid \boldsymbol{\xi}_i, \boldsymbol{\theta}) P(\boldsymbol{\xi}_i \mid \boldsymbol{\theta})
}\end{equation}

With this we can compute the probability for each hidden state and
subsequently we can compute the probability for each LNL \(v\) to be
involved given the diagnosis \(\mathbf{Z}\) and the model parameters
\(\boldsymbol{\theta}\) by summing over all possible hidden states
\(\boldsymbol{\xi}_{i,v}\) where the LNL \(v\) is involved:

\begin{equation}\phantomsection\label{eq-risk}{
P(X_v=1 \mid \mathbf{Z}, \boldsymbol{\theta}) = \sum_{\boldsymbol{\xi}_{i,v}} P(\mathbf{X} =\boldsymbol{\xi}_i \mid \mathbf{Z}, \boldsymbol{\theta})
}\end{equation}

As opposed to the model training, we do not assume anymore that the
diagnoses correspond to the hidden states. Instead, we consider the
sensitivity and specificity of the diagnostic modalities used to
determine the involvement of LNLs. Enabling the uncertainty originating
from diagnostic modalities to be considered in the model. Here we chose
the sensitivity of 0.81 for clinical modalities corresponding to the
literature value for imaging modalities \citep{de_bondt_detection_2007}
and specificity of 1.0, indicating that all positively diagnosed LNLs
are indeed involved.

@@

\begin{itemize}
\tightlist
\item
  \(\pi_m^{s_k}\) is the mixing coefficient for subsite \(s_k\) (where
  patient \(k\) has their tumor) and model \(m\),
\item
  \(P(\mathbf{D}_k \mid \boldsymbol{\theta}_m)\) is the likelihood of
  patient \(k\)'s diagnosis, i.e.~The involvement state
  \(\mathbf{X}_k = \boldsymbol{\xi}_i\) and t-stage \(T_k\), given that
  it was generated by model \(m\),
\item
  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
\item
  \(\epsilon_{k}^m\) is the latent variable that indicates patient \(k\)
  was generated by model \(m\).
\end{itemize}

@@

To find the mixing coefficients \(\boldsymbol{\pi}\) and all models'
parameters \(\boldsymbol{\theta}_m\) that maximize
equation~\ref{eq-incomplete-likelihood}, we follow an iterative approach
called \emph{expectation-maximization} or \emph{EM-algorithm}. With
arbitrarily initialized starting parameters, we alternate between the
following two steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In the \textbf{E}xpectation step, we compute the responsibilities
  \(\gamma(\epsilon_k^m)\), which represent the probabilities of a
  patient \(k\) originating from one of the models \(m\), given the
  current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
  as given in equation~\ref{eq-responsibilities}.
\item
  During the \textbf{M}aximization step, we find a new set of parameters
  that maximize the new expected complete data likelihood
  (equation~\ref{eq-expected-complete-likelihood}). For the mixture
  coefficients, we can even find an analytic solution to the new
  maximum: \[
  \pi_m^s = \frac{1}{|K_s|} \sum_{k \in K_s} \gamma(\epsilon_k^m)
  \] where we sum over the set of all patients \(K_s\) with their tumor
  in subsite \(s\).\\
  The models' new parameters are found by numerically maximizing the
  respective likelihood, weighted by the responsibilities:
  \begin{equation}\phantomsection\label{eq-maximization-step}{
  \ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_k^K \gamma(\epsilon^m_{k}) \big[ \ln \pi_m^{s_k} + \ln P \left( \mathbf{D}_k \mid \boldsymbol{\theta}_m \right) \big]
  }\end{equation}
\end{enumerate}

@@

\subsection{Uncertainty estimation}\label{sec-uncert}

To estimate the uncertainty associated with both the model parameters
and the resulting risk estimates, we employ a bootstrapping approach.
Although more comprehensive sampling methods exist for the EM
algorithm---such as the imputation-posterior (IP) algorithm, which
samples from both model parameters and latent variables---we opt for a
computationally more efficient alternative that still yields multiple
estimates of all parameters.

Our resampling procedure involves the following steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Resample the observed data with replacement, maintaining the original
  number of samples per ICD code.
\item
  Fit the model to each resampled dataset.
\item
  Extract the model parameters and risk estimates from each fitted
  model.
\end{enumerate}

This method enables a straightforward uncertainty analysis while
avoiding the computational cost of more intensive sampling techniques.

\section{Model Training Evaluation}\label{sec-model_train}

In the following sections we will analyze the model training for three
components first, considering Oral Cavity, Oropharynx and Hypopharynx
data. Then we extend the model to four components, additionally
including Larynx data. Both models will include LNLs I, II, III, IV and
V. LNLs VI and VII are omitted for simplicity and since the prevalence
for these LNLs never exceeds 11\% (figure~\ref{fig-involvement}),
meaning that the risk of involvement is low in most cases.

\subsection{Three Component Mixture Model}\label{sec-3comp}

We first evualuate the methodology for a mixture model with M = 3
components. We include the ICD codes as subsites for oral cavity,
hypopharynx and oropharynx. In figure~\ref{fig-convergence} the
convergence of the negative log-likelihood and change in model
parameters is depicted. After a random inizialization, the algorithm
rapidly converges. The algorithm was stopped when the difference of
log-likelihood between two iterations was below 0.01.

@@

The spatial plot in figure~\ref{fig-3_simplex} illustrates how the model
assigns the three components to different tumor subsites. Component 0,
located at the bottom right of the triangle, primarily characterizes
hypopharyngeal subsites. Both hypopharyngeal subsites, piriform sinus
(C12) and hypopharynx (C13), are strongly assigned to this component.
Similarly, component 1, located at the bottom left, characterizes oral
cavity subsites. For instance, the gum (C03), which has the highest LNL
I involvement is fully assigned to this subsite. As subsites
anatomically approach the oropharynx, their mixture coefficients for the
oropharynx-like component, at the top of the triangle, increase. This is
evident in the subsites C02 (tongue) and C05 (palate), which display a
higher proportion of oropharyngeal influence in their mixture. These
results conform well with the involvement patterns observed in the data
in figure figure~\ref{fig-involvement}. The base of tongue subsite
(C01), which exhibits the highest involvement of LNL II, is fully
assigned to oropharynx-like component. Similarly, subsite C10, which
includes several oropharyngeal regions, is assigned roughly 50\% to the
oropharynx-like component, with the remaining mixture distributed across
the other two components, representing its large variation in spread
patterns.

@@

\includegraphics{figures/mixture_components_3_simplex.png}

@@

\includegraphics{figures/mixture_components_3.png}

@@

To evaluate the uncertainty in the model parameters, the bootstrapping
approach introduced in section~\ref{sec-uncert} has been applied.
Specifically, we generated 200 bootstrap samples by resampling the
observed data with replacement, maintaining the original number of
samples per ICD code. For each bootstrap sample, the mixture model was
re-fitted, and the resulting parameters were recorded. This process
yielded a distribution of parameter estimates, allowing us to compute
confidence intervals and assess the variability in the model's
predictions.

The results of the bootstrapping analysis are shown in
figure~\ref{fig-simplex_uncert} and
figure~\ref{fig-3_comp_scatter_uncert}. The figure illustrates the
variability in the mixture coefficients \(\boldsymbol{\pi}\) for each
subsite in the same representation as figure~\ref{fig-3_simplex}. The
figure shows that the uncertainties match the continuity of the spread
patterns. The base of tongue (C01) has most of its uncertainty along in
between the oropharynx component and the hypopharynx component.
Similarly, the hypopharynx subsite (C13) mostly presents uncertainty
between these two components. The more general oropharnyx subsite (C10)
shows more uncertainty in each components direction, due to its mixed
spread pattern. The subsite for other pars of mouth (C03) however, only
has uncertainty along the axis connecting the oropharyngeal component
and the oral cavity componen, therefore never mixing with the
hypopharynx-like component on the right. In figure
figure~\ref{fig-3_comp_scatter_uncert} the asymmetry of the uncertainty
is better depicted. E.g. the base of tongue (C01) subsite and the
hypopharynx (C13) subsite have strongly asymmetric uncertainties.
Further, the plot better characterizes the uncertainty in the palate
subsite (C05), where the mixture and uncertainty is mainly in the
oropharynx-like and the oral cavity-like components if we center the
uncertainty around the optimal values instead of the mean values as it
is depicted in figure~\ref{fig-simplex_uncert}.

\includegraphics{figures/mixture_coefficients_uncert_simplex.png}

\caption{\label{fig-simplex_uncert}Variability in mixture coefficients
\(\boldsymbol{\pi}\) across bootstrap samples for each subsite. Where
the dots represent the optimal values and the ellipses represent the
68\% percentile centered around the mean value. The 68\% percentile are
computed assuming a 2D-gaussian distribution. However, with the bounds
of 0 and 1 the sampled mixture coefficients do not exactly follow a
gaussian distribution.}

\includegraphics{figures/3_comp_mixture_coefficients_uncert_scatter.png}

\caption{\label{fig-3_comp_scatter_uncert}Variability in mixture
coefficients \(\boldsymbol{\pi}\) across bootstrap samples for each
subsite. The dots represent the optimal values. The bars include 68\% of
all sample values centered around the optimal values.}

Additionally, the bootstrapping approach enables us to quantify the
uncertainty in the model predictions for each subsite and LNL. To
benchmark the model we analyze the predicted and observed prevalences
for LNLs I, II, III, IV, and V. These predictions are stratified by
tumor subsite and t-stage, providing a comprehensive view of the model's
performance and its associated uncertainty.

In figure~\ref{fig-oral_cavity} we present the predictions for the oral
cavity subsites C03 and C05. Here we can clearly see in figure
figure~\ref{fig-oral_cavity_C03_early} and
figure~\ref{fig-oral_cavity_C03_late} that the difference in prevalence
is substantial, even though both subsites are grouped in the same tumor
location. The model predicts a higher prevalence for LNL II in subsite
C05 (palate) compared to subsite C03 (gum). This is consistent with the
observed data, where the prevalence of LNL II involvement is higher for
the palate than for the gum. However, the model only predicts a slightly
higher prevalence for LNL I in subsite C03 compared to subsite C05. This
is not consistent with the observed data, where the prevalence of LNL I
involvement is significantly higher for the gum than for the palate.
Thus, showing a limitation in the model to properly split these
different involvements. In late t-stages, depicted in figure
figure~\ref{fig-oral_cavity_C03_late} and
figure~\ref{fig-oral_cavity_C05_late}, the model predicts a higher
prevalence for LNL II in subsite C05 compared to the early t-stage, but
not fully capturing the increase in prevalence. For subsite C03 the LNL
II involvement is predicted to increase as well, capturing the increase
in prevalence. However, for LNL I, while capturing the increase in
prevalence, it does not optimally predict the prevalence, similarly to
the early t-stage. For LNLs III, IV and V, the model consistently
predicts the low prevalences observed in the data for both subsites.

\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C03_early.png}

\subcaption{\label{fig-oral_cavity_C03_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C05_early.png}

\subcaption{\label{fig-oral_cavity_C05_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C03_late.png}

\subcaption{\label{fig-oral_cavity_C03_late}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C05_late.png}

\subcaption{\label{fig-oral_cavity_C05_late}}

\caption{\label{fig-oral_cavity}Observed prevalence (solid lines) and
predicted prevalence (histograms) for subsites C03 and C05. Figures a
and b show the prevalences for early t-stages. Figures c and d show the
prevalences for late t-stages.}

In figure~\ref{fig-oropharynx} we present the predictions for the
oropharynx subsites C01 and C10. In figures
figure~\ref{fig-oropharynx_C01_early} and
figure~\ref{fig-oropharynx_C10_early} we can see that the model predicts
a higher prevalence for LNL II in subsite C01 (base of tongue) compared
to subsite C10 (oropharynx). This is consistent with the observed data.
In figures figure~\ref{fig-oropharynx_C01_late} and
figure~\ref{fig-oropharynx_C10_late} we can see that the model precisely
captures the increased prevalence in late t-stage for both subsites.
LNLs I, IV and V are predicted to have low prevalences for both
subsites, but the model does not perfectly fit the predicted prevalences
to the observed prevalences.

\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C01_early.png}

\subcaption{\label{fig-oropharynx_C01_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C10_early.png}

\subcaption{\label{fig-oropharynx_C10_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C01_late.png}

\subcaption{\label{fig-oropharynx_C01_late}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/3_predicted_prevalence_C10_late.png}

\subcaption{\label{fig-oropharynx_C10_late}}

\caption{\label{fig-oropharynx}Observed prevalence (solid lines) and
predicted prevalence (histograms) for subsites C01 and C10. Figures a
and b show the prevalences for early t-stages. Figures c and d show the
prevalences for late t-stages.}

In summary, the mixture model successfully distinguishes between
subsites, capturing the continuous variation in lymphatic spread
patterns across tumor locations. However, the model's predictions are
not always perfect. While it generally aligns with observed data,
certain nuances, such as the higher prevalence of LNL I involvement in
subsite C03 compared to C05, are not fully captured.

\subsection{Four Component Mixture Model}\label{sec-4comp}

@@

The component assignment is shown in figure~\ref{fig-4-matrix}.
Similarly to the 3-component model the different tumor locations are
assigned to a one of the components. In this case, the new component is
used to model laryngeal subsites. The uncertainty of the assignments is
shown in figure~\ref{fig-4_comp_scatter_uncert}. The most striking
differences are in subsites C03 and C12, which were formerly assigned to
a single component. Now, C03 mixes with the laryngeal component, while
C12 is mixed with the laryngeal and the hypopharyngeal component.
Another interesting observation is the mixture of the supraglottis
subsite (C32.1) which is anatomically close to the hypopharynx. The
model assigns a strong mixture of the larynx-like component and the
hypopharynx-like component to this subsite. This is consistent with the
observed data, where the supraglottis subsite has a high prevalence of
LNL II and III involvement, which is also observed in the hypopharynx
subsites. In contrast, the glottis subsite (C32.0) is strongly assigned
to the laryngeal component, while the subsite C32.2 (subglottis) is
assigned to the laryngeal component with little uncertainty. This is
consistent with the observed data, where the glottis subsite has a much
lower prevalence of involvement in any LNL. However, there is a strong
t-stage dependency for this subsite.

\begin{minipage}{0.29\linewidth}

\includegraphics{figures/mixture_components_4.png}

\caption{\label{fig-4-matrix}Matrix representation of component
assignment. Each row of the matrix corresponds to each ICD code. The
columns represent the four different components}

\end{minipage}%
%
\begin{minipage}{0.71\linewidth}

\includegraphics{figures/4_comp_mixture_coefficients_uncert_scatter.png}

\caption{\label{fig-4_comp_scatter_uncert}Assignment of each subsite to
each of the four components. The closer a subsite is to a vertex, the
more it is assigned to the corresponding component, with component 0 on
the bottom right, 1 on the bottom left, 2 on the top left and 3 on the
top right. The size of the marker (area) corresponds to the number of
patients in each subsite.}

Similarly to the three component model, we can analyze the predicted and
observed prevalences for LNLs I, II, III, IV and V. We will first
revisit the oral cavity subsites C03 and C05. to check, whether the
additional component improves the model predictions. In
figure~\ref{fig-oral_cavity_4} the predictions for the oral cavity
subsites C03 and C05 are shown. The model now predicts a higher
prevalence for LNL I in subsite C03 (gum) compared to subsite C05
(palate) for both early and

\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C03_early.png}

\subcaption{\label{fig-oral_cavity_C03_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C05_early.png}

\subcaption{\label{fig-oral_cavity_C05_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C03_late.png}

\subcaption{\label{fig-oral_cavity_C03_late}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C05_late.png}

\subcaption{\label{fig-oral_cavity_C05_late}}

\caption{\label{fig-oral_cavity_4}Observed prevalence (solid lines) and
predicted prevalence (histograms) for subsites C03 and C05. Figures a
and b show the prevalences for early t-stages. Figures c and d show the
prevalences for late t-stages.}

The predicted prevalences for for the laryngeal subsites C32.0 and C32.1
are shown in figure~\ref{fig-larynx_4}. Here, the large difference
between early and late t-stages is clearly visible. For subsite C32.0
(glottis) the model predicts a very low prevalence for LNL I, II and III
in early t-stages, which is consistent with the observed data. For late
t-stages, the model captures the increased involvement probabilities for
LNLs II and III. For subsite C32.1 (supraglottis) the model
overestimates the prevalence for LNLs II and III in early t-stages and
slightly underestimates their prevalence in late t-stages.

\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C32.0_early.png}

\subcaption{\label{fig-oral_cavity_C32.0_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C32.1_early.png}

\subcaption{\label{fig-oral_cavity_C05_early}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C32.0_late.png}

\subcaption{\label{fig-oral_cavity_C03_late}}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\includegraphics{figures/4_predicted_prevalence_C32.1_late.png}

\subcaption{\label{fig-oral_cavity_C05_late}}

\caption{\label{fig-larynx_4}Observed prevalence (solid lines) and
predicted prevalence (histograms) for subsites C32.0 and C32.1. Figures
a and b show the prevalences for early t-stages. Figures c and d show
the prevalences for late t-stages.}

The model parameters for the larynx-like component are shown in
figure~\ref{fig-4_larynx_parameters}. The parameter that governs the
impact of t-stage \(p_late\) is estimated around 1. Therefore, the model
maximizes the increase of risk of involvement for late t-stages to
capture the large difference in prevalence between early and late
t-stages.

@@

\includegraphics{figures/larynx_component_values.png}

@@

\caption{\label{fig-4_larynx_parameters}Model parameters for the
larynx-like component. The optimal value and the 68\% percentile around
the optimal value are marked in red. The mean values and extreme values
are indicated by the black bars.}

@@

\section{Model Predictions}\label{sec-model_pred}

Given the model parameters, we can compute the risk of occult metastases
given a patient's diagnosis. The model can be applied to predict the
risk of involvement for each LNL, given the patient's t-stage and
primary tumor subsite.

The risk predictions are computed by applying the equation~\ref{eq-risk}
to each component and subsequently mixing the results using the mixture
coefficients \(\boldsymbol{\pi}\). The risk of involvement for each LNL
\(v\) in subsite \(s\) is given by:
\begin{equation}\phantomsection\label{eq-risk-mixing}{
P(X_v^s=1 \mid \mathbf{D}, \boldsymbol{\theta}) = \sum_{m=1}^M \pi_m^{s} P(X_v=1 \mid \mathbf{D}, \boldsymbol{\theta}_m)
}\end{equation}

where \(s\) is the subsite and \(\boldsymbol{\theta}_m\) are the
parameters of model component \(m\).

For all estimates we will set a 5\% threshold for the risk of
involvement. This means that if the model predicts a risk of involvement
of 5\% or higher, we will consider the LNL to be at a ``high'' risk of
involvement and should be irradiated. Consequently all LNLs with a risk
of involvement below 5\% will be considered to be at a ``low'' risk of
involvement and should not be irradiated. This allows us to compare the
model predictions with the current clinical guidelines for elective
irradiation of LNLs.

In the following sections we will analyze the model predictions for the
four component model only and for selected clinical diagnoses.

\subsection{Oral cavity}\label{oral-cavity}

Oral cavity tumors typically spread to LNLs I, II and III as shown in
figure~\ref{fig-involvement}. Consequently, the clinical guidelines
recommend irradiating LNLs I, II, III for all tumors, irrespective of
t-stage and even if there is no LNL involvement
\citep{biau_selection_2019}.

We first analyze the risk predictions for the case where we diagnosed

\section{Appendix}\label{appendix}

Here I will probably add more figures such as all component parameter
estimates and their uncertainties.
